The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-23 degruyter-crelle/dictionary.tex
Moved the current version into the crelle folder. Work there.

% !TEX encoding = UTF-8 Unicode
\documentclass{amsart}
\pdfoutput=1

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%

@@

\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{function--sheaf dictionary, trace of Frobenius, geometrization, categorification, quasicharacters, quasicharacter sheaves, smooth commutative group schemes, finite fields, local fields, algebraic tori, Neron models, Greenberg functor}

@@

%%%%%%%%%%%% BEGIN DOCUMENT %%%%%%%%%%%%
\begin{document}

@@

\maketitle

@@

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over $\FF_q$, then the trace of Frobenius provides a perfect dictionary between characters of $G(\FF_q)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \qcs{E} \iso \qcs{E} \boxtimes \qcs{E},\tag{0}
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider sheaf homorphisms $\qcs{E} \to \qcs{E}'$ which are compatible with particular choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E'}$. 
{\it A priori}, these sheaf homomorphisms do not form a category, but since $G$ is connected, the existence of \eqref{introbox} makes \eqref{introbox} unique; as a consequence, there is no impediment to categorifying the dictionary in this case.

If $G$ is a commutative algebraic group over $\FF_q$ which is not connected, however, then one must keep track of the choice of \eqref{introbox}, leading to a category of pairs $(\qcs{E},\mu_\qcs{E})$ where $\qcs{E}$ is a rank-one local system on $G$ and $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ is an isomorphism of local systems on $G\times G$. 
If the component group scheme $\pi_0(G)$ (which is an étale commutative group scheme over $\FF_q$) is not cyclic, then while the trace of Frobenius provides a surjection from isomorphism classes in the category of pairs $(\qcs{E},\mu_\qcs{E})$ to characters of $G(\FF_q)$, that group homomorphism need not be injective. 
In other words, if $G$ is a commutative algebraic group over $\FF_q$ which is not connected, then every character may be geometrized as a pair $(\qcs{E},\mu_\qcs{E})$ but perhaps not uniquely. 
While this might appear to be a defect in the dictionary for characters of commutative algebraic groups over finite fields, it is not serious since if $(\qcs{E},\mu_\qcs{E})$ and $(\qcs{E}',\mu_\qcs{E}')$ determine the same character of $G(\FF_q)$ then $\qcs{E}\iso \qcs{E}'$ as local systems on $G$.

@@

We first test the category of quasicharacter sheaves by examining its objects on étale commutative group schemes $G$ over $\FF_q$; étale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an étale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, canonically.
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ is not an isomorphism should not be too troubling because if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; for example, without them, we would not be able to geometrize the character of the étale group scheme $\ZZ$ over $\FF_q$ given by $1\mapsto \ell$, for example.

Returning to the case of connected commutative algebraic groups $G$ over $\FF_q$, we confirm that, in this case, $\QC(G)$ is equivalent to the category of characters $\psi$ of $G(\FF_q)$, the equivalence being given by descent to $G$ followed by pullback by the Lang isogeny for $G$. 
%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ and the trace of Frobenius gives an isomorphism $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\FF_q$, then $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ measures the kernel of $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$. 
%This observation makes plain the fact that  . . .
%
As we show in the paper, while characters of connected commutative algebraic groups $G$ can be geometrized using  local systems trivialized by discrete isogenies to $G$, this is not the right perspective if one wishes to geometrize characters of smooth commutative group schemes, such as étale commutative group schemes. Indeed, quasicharacter sheaves on étale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood quasicharacter sheaves on commutative smooth group schemes over $\FF_q$ in two extreme cases -- connected algebraic groups and étale commutative group schemes -- we turn to the general case. Remarkably, we find that the general case can be understood from these two extreme cases, using the component group sequence
\begin{equation}
0 \to G^0 \to G \to \pi_0(G) \to 0.
\end{equation}
The main result of the paper uses this sequence to find $\QCiso{G}$ by situating it in a canonical short exact sequence
\begin{equation}
0 \to \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \to \QCiso{G} \to \Hom(G(\FF_q),\EEx) \to 0.
\end{equation}
We remark that $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$ is non-trivial even for some commutative algebraic groups.
We also illuminate the nature of the category $\QC(G)$ by showing that every morphism in this category is either an isomorphism or trivial, and by showing that 
\begin{equation}
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx),
\end{equation}
canonically, for all quasicharacter sheaves $\qcs{L}$ on $G$.

@@

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(X_K)$ of quasicharacter sheaves for $X_K$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the Néron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(\FF_q), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

In Section~\ref{sec:main} we prove that the rows of \eqref{eq:intro-snake} are exact and
use the snake lemma to determine the kernel and cokernel of $\TrFrob{G}$ for general $G$.
We show that $\QCiso{G^0} \to \QCiso{G}$ is surjective in Section~\ref{ssec:restriction} and
that the remainder of the top row of \eqref{eq:intro-snake} is exact in Section~\ref{ssec:component}.
We put it together in Theorem~\ref{thm:snake} of Section~\ref{ssec:snake}, showing that $\TrFrob{G}$
is surjective with kernel $\Hh^2(\pi_0(\bG), \EEx)^{\Frob{}}$.  Having described the isomorphism classes of objects,
in Theorem~\ref{thm:autornaught} of Section~\ref{ssec:qcmor} we show that
\[
\Aut(\qcs{L}) \cong \Hom(\pi_0(\bG)_{\Weil{}}, \EEx)
\]
and that every morphism is either trivial or an isomorphism.  We conclude in Section~\ref{sec:basechange} by
giving functorial versions of the restriction and norm homomorphisms $G(k')^* \to G(k)^*$ and $G(k)^* \to G(k')^*$
and describing how quasicharacter sheaves behave under Weil restriction.

@@

%\begin{acknowledgements}
\subsection{Acknowledgments}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We thank Takashi Suzuki for some very helpful observations and suggestions, including those related to our use of the Hochschild-Serre spectral sequence.
We thank Joseph Bernstein for suggesting a connection between our category of quasicharacter sheaves on $G$ and his sheaves on the stacky classifying space of $G$.
Finally, we gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our week-long stay in May 2012.

%\end{acknowledgements}

\tableofcontents

@@

%\'Etale group schemes form a counterpoint to connected groups,
%since the component group of a smooth commutative group scheme
%is an \'etale group scheme \cite{vdGeer-Moonen:AbelianVarieties}*{III, \S 4}.
%However, the function--sheaf dictionary is substantially different in the two cases.
%To see why, we begin by specializing the definition of quasicharacter sheaves to the case
%of étale commutative group schemes,
%using the equivalence $G \mapsto G(\bFq)$
%from the category of \'etale group schemes over $\Fq$ to the category of groups equipped
%with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.

In this section we find a complete characterization of the category of quasicharacter sheaves on étale commutative group schemes over finite fields.

@@

The equivalence $G \mapsto G(\bFq)$
from the category of \'etale commutative group schemes over $\Fq$ to the category of commutative groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group,
provides the following simple description of quasicharacter sheaves.
%
A quasicharacter sheaf $\qcs{L}$ on an étale commutative group scheme $G$ over $\Fq$ is:
\begin{enumerate}
 \labitem{(qc.1)}{qc.1} an indexed set of one-dimensional
  $\EE$-vector spaces $\gqcs{L}_x$, as $x$ runs over
  $G(\bFq)$;

@@

\begin{remark}
Although we will only use the group homomorphism $S_G : \QCiso{G}\to \Hh^2(E^\bullet_G)$ when $G$ is an étale commutative group scheme, the function is defined for all smooth commutative group schemes $G$.
However, the next result is false in that context.
\end{remark}

@@

In this section we find the relationship between the group homomorphisms $\TrFrob{G}$ and $S_G$, and use this to find the group $\QCiso{G}$ when $G$ is an étale commutative group scheme over $\Fq$.

@@

\begin{proposition}\label{prop:autornaught_etale}
Let $G$ be an étale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

%Note that Proposition~\ref{prop:autornaught_etale} tells us that if $G$ is an étale commutative group scheme over $\Fq$ then every hom-set is either trivial or $\{ 0 \} \cup \left(G(\bFq)_{\Weil{}}\right)^*$, non-canonically.

@@

A finite, étale, surjective morphism $H\to G$ of smooth group schemes over $\Fq$ for which the action of $\Gal(\bFq/\Fq)$ on the kernel is trivial is called a {\it discrete isogeny}.

@@

\begin{proof}
Let $f$, $A$, $V$ and $\psi$ be as above; set $\qcs{E} = (f_! V_H)_\psi$.
It is clear that $\qcs{E}$ is an $\ell$-adic local system on $G$ of rank one.
We must show that comes equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$.
To do this we use étale descent to see that pullback along $f$ gives an equivalence between $\ell$-adic local systems on $G$ and $A$-equivariant local systems on $H$. 
In particular, $f^*\qcs{E}$ is the $A$-equivariant constant sheaf $V$ on $H$ with character $\psi$.
Since $f$ is a morphism of group schemes, the functor $f^*$ defines $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$
from the isomorphism $m^*\psi \iso \psi \boxtimes\psi$ determined by $V\to V\otimes V$.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by étale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\subsection{Quasicharacter sheaves on the Greenberg transform of a Néron model} \label{ssec:qc_on_GN}

>>>>

%%------------------------------------------------------%%
%% This file should be the main file of your article.   %%
%% Please uncomment exactly one of the following lines. %%
%%------------------------------------------------------%%

%% \documentclass[CM,Submssn]{degruyter-crelle}       %% Equations numbered as (1), (2) etc.
 \documentclass[CM,Submssn,SecEq]{degruyter-crelle} %% Equations numbered as (1.1), (1.2) etc.

%% Do not alter the following line.
\firstpage{}\volume{}\volumeyear{}\copyrightyear{}\doiyear{}\doi{}\received{XXX}\revised{XXX}

@@

\author{Clifton}{Cunningham}{C.~Cunningham}{Calgary}
\author{David}{Roe}{D.~Roe}{Calgary}

\contact{Department of Mathematics and Statistics, University of Calgary, street, zip code, city, country}{cunning@math.ucalgary.ca}
\contact{Pacific Institute for the Mathematical Sciences, University of Calgary, street, zip code, city, country}{roed.math@gmail.com}

%\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
%\keywords{function--sheaf dictionary, trace of Frobenius, geometrization, categorification, quasicharacters, quasicharacter sheaves, smooth commutative group schemes, finite fields, local fields, algebraic tori, Neron models, Greenberg functor}

\researchsupported{We gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our week-long stay in May 2012.}

%% Define theorem-like environments as usual:

%\theoremstyle{plain}
 %\newtheorem{theorem}{Theorem}[section]

%\theoremstyle{definition}
% \newtheorem{remark}[theorem]{Remark}

@@

%%%%%%%%%%%% BEGIN DOCUMENT %%%%%%%%%%%

\begin{document}

@@

%% \tableofcontents %% Just for papers exceeding 50 pages.

@@

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over $\FF_q$, then the trace of Frobenius provides a perfect dictionary between characters of $G(\FF_q)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \qcs{E} \iso \qcs{E} \boxtimes \qcs{E},%\tag{0}
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider sheaf homorphisms $\qcs{E} \to \qcs{E}'$ which are compatible with particular choices of \eqref{introbox} for $\qcs{E}$ and $\qcs{E'}$. 
{\it A priori}, these sheaf homomorphisms do not form a category, but since $G$ is connected, the existence of \eqref{introbox} makes \eqref{introbox} unique; as a consequence, there is no impediment to categorifying the dictionary in this case.

If $G$ is a commutative algebraic group over $\FF_q$ which is not connected, however, then one must keep track of the choice of \eqref{introbox}, leading to a category of pairs $(\qcs{E},\mu_\qcs{E})$ where $\qcs{E}$ is a rank-one local system on $G$ and $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ is an isomorphism of local systems on $G\times G$. 
If the component group scheme $\pi_0(G)$ (which is an \'etale commutative group scheme over $\FF_q$) is not cyclic, then while the trace of Frobenius provides a surjection from isomorphism classes in the category of pairs $(\qcs{E},\mu_\qcs{E})$ to characters of $G(\FF_q)$, that group homomorphism need not be injective. 
In other words, if $G$ is a commutative algebraic group over $\FF_q$ which is not connected, then every character may be geometrized as a pair $(\qcs{E},\mu_\qcs{E})$ but perhaps not uniquely. 
While this might appear to be a defect in the dictionary for characters of commutative algebraic groups over finite fields, it is not serious since if $(\qcs{E},\mu_\qcs{E})$ and $(\qcs{E}',\mu_\qcs{E}')$ determine the same character of $G(\FF_q)$ then $\qcs{E}\iso \qcs{E}'$ as local systems on $G$.

@@

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$; \'etale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an \'etale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, canonically.
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ is not an isomorphism should not be too troubling because if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; for example, without them, we would not be able to geometrize the character of the \'etale group scheme $\ZZ$ over $\FF_q$ given by $1\mapsto \ell$, for example.

Returning to the case of connected commutative algebraic groups $G$ over $\FF_q$, we confirm that, in this case, $\QC(G)$ is equivalent to the category of characters $\psi$ of $G(\FF_q)$, the equivalence being given by descent to $G$ followed by pullback by the Lang isogeny for $G$. 
%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ and the trace of Frobenius gives an isomorphism $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\FF_q$, then $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ measures the kernel of $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$. 
%This observation makes plain the fact that  . . .
%
As we show in the paper, while characters of connected commutative algebraic groups $G$ can be geometrized using  local systems trivialized by discrete isogenies to $G$, this is not the right perspective if one wishes to geometrize characters of smooth commutative group schemes, such as \'etale commutative group schemes. Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood quasicharacter sheaves on commutative smooth group schemes over $\FF_q$ in two extreme cases -- connected algebraic groups and \'etale commutative group schemes -- we turn to the general case. Remarkably, we find that the general case can be understood from these two extreme cases, using the component group sequence
\begin{equation}
0 \to G^0 \to G \to \pi_0(G) \to 0.
\end{equation}
The main result of the paper uses this sequence to find $\QCiso{G}$ by situating it in a canonical short exact sequence
\begin{equation}
0 \to \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \to \QCiso{G} \to \Hom(G(\FF_q),\EEx) \to 0.
\end{equation}
We remark that $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$ is non-trivial even for some commutative algebraic groups.
We also illuminate the nature of the category $\QC(G)$ by showing that every morphism in this category is either an isomorphism or trivial, and by showing that 
\begin{equation}
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx),
\end{equation}
canonically, for all quasicharacter sheaves $\qcs{L}$ on $G$.

@@

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(X_K)$ of quasicharacter sheaves for $X_K$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the N\'eron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(\FF_q), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \Hom(\pi_0(G)(\FF_q),\EEx) \rar & \Hom(G(\Fq),\EEx) \rar & \Hom(G^0(\FF_q),\EEx) \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

In Section~\ref{sec:main} we prove that the rows of \eqref{eq:intro-snake} are exact and
use the snake lemma to determine the kernel and cokernel of $\TrFrob{G}$ for general $G$.
We show that $\QCiso{G^0} \to \QCiso{G}$ is surjective in Section~\ref{ssec:restriction} and
that the remainder of the top row of \eqref{eq:intro-snake} is exact in Section~\ref{ssec:component}.
We put it together in Theorem~\ref{thm:snake} of Section~\ref{ssec:snake}, showing that $\TrFrob{G}$
is surjective with kernel $\Hh^2(\pi_0(\bG), \EEx)^{\Frob{}}$.  Having described the isomorphism classes of objects,
in Theorem~\ref{thm:autornaught} of Section~\ref{ssec:qcmor} we show that
\[
\Aut(\qcs{L}) \cong \Hom(\pi_0(\bG)_{\Weil{}}, \EEx)
\]
and that every morphism is either trivial or an isomorphism.  We conclude in Section~\ref{sec:basechange} by
giving functorial versions of the restriction and norm homomorphisms $\Hom(G(k'),\EEx) \to \Hom(G(k),\EEx)$ and $\Hom(G(k),\EEx) \to \Hom(G(k'),\EEx)$
and describing how quasicharacter sheaves behave under Weil restriction.

@@

%\tableofcontents

@@

%\'Etale group schemes form a counterpoint to connected groups,
%since the component group of a smooth commutative group scheme
%is an \'etale group scheme \cite{vdGeer-Moonen:AbelianVarieties}*{III, \S 4}.
%However, the function--sheaf dictionary is substantially different in the two cases.
%To see why, we begin by specializing the definition of quasicharacter sheaves to the case
%of \'etale commutative group schemes,
%using the equivalence $G \mapsto G(\bFq)$
%from the category of \'etale group schemes over $\Fq$ to the category of groups equipped
%with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group.

In this section we find a complete characterization of the category of quasicharacter sheaves on \'etale commutative group schemes over finite fields.

@@

The equivalence $G \mapsto G(\bFq)$
from the category of \'etale commutative group schemes over $\Fq$ to the category of commutative groups equipped
with an action of $\Gal(\bFq/\Fq)$, continuous for the discrete topology on the group,
provides the following simple description of quasicharacter sheaves.
%
A quasicharacter sheaf $\qcs{L}$ on an \'etale commutative group scheme $G$ over $\Fq$ is:
\begin{enumerate}
 \labitem{(qc.1)}{qc.1} an indexed set of one-dimensional
  $\EE$-vector spaces $\gqcs{L}_x$, as $x$ runs over
  $G(\bFq)$;

@@

\begin{remark}
Although we will only use the group homomorphism $S_G : \QCiso{G}\to \Hh^2(E^\bullet_G)$ when $G$ is an \'etale commutative group scheme, the function is defined for all smooth commutative group schemes $G$.
However, the next result is false in that context.
\end{remark}

@@

In this section we find the relationship between the group homomorphisms $\TrFrob{G}$ and $S_G$, and use this to find the group $\QCiso{G}$ when $G$ is an \'etale commutative group scheme over $\Fq$.

@@

\begin{proposition}\label{prop:autornaught_etale}
Let $G$ be an \'etale commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. Moreover, the trace map induces an isomorphism of groups
\[
\Aut(\qcs{L}) \to \Hom(G(\bFq)_{\Weil{}}, \EEx).
\]
\end{proposition}

@@

%Note that Proposition~\ref{prop:autornaught_etale} tells us that if $G$ is an \'etale commutative group scheme over $\Fq$ then every hom-set is either trivial or $\{ 0 \} \cup \left(G(\bFq)_{\Weil{}}\right)^*$, non-canonically.

@@

A finite, \'etale, surjective morphism $H\to G$ of smooth group schemes over $\Fq$ for which the action of $\Gal(\bFq/\Fq)$ on the kernel is trivial is called a {\it discrete isogeny}.

@@

\begin{proof}
Let $f$, $A$, $V$ and $\psi$ be as above; set $\qcs{E} = (f_! V_H)_\psi$.
It is clear that $\qcs{E}$ is an $\ell$-adic local system on $G$ of rank one.
We must show that comes equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$.
To do this we use \'etale descent to see that pullback along $f$ gives an equivalence between $\ell$-adic local systems on $G$ and $A$-equivariant local systems on $H$. 
In particular, $f^*\qcs{E}$ is the $A$-equivariant constant sheaf $V$ on $H$ with character $\psi$.
Since $f$ is a morphism of group schemes, the functor $f^*$ defines $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$
from the isomorphism $m^*\psi \iso \psi \boxtimes\psi$ determined by $V\to V\otimes V$.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\subsection{Quasicharacter sheaves on the Greenberg transform of a N\'eron model} \label{ssec:qc_on_GN}
